---
title: "Foraminifera community composition"
author: "Venus van Hoof"
output:
  html_document:
    df_print: paged
---
This script contains code for visualizations on the identfied species, community compostion and calculation for the shannon diversity and ASV richness.

**Identified species**


In the script below, a barplot is made, showing which species were identified (> 99.4% ID) and the number of reads for each species is shown.

```{r}
setwd("C:/Users/Venus/OneDrive/Documents/Github/RajaAmpat_forams")

library(tidyverse)
library(extrafont)
library(fitdistrplus)
library(ggh4x)
library(car)
library(FSA)

df <- read.csv("Rdata/working_dataset.csv")
sum(df$reads)
foram_df <- subset(df, df$Phylum == "Foraminifera")

#check the read and ASV abundance with higher taxonomic resolution
sum(foram_df$reads) #5708287 reads
length(unique(foram_df$otu)) #2980 otus
foram_ord <-  foram_df[foram_df$X.Identity.percentage > 84,]
sum(foram_ord$reads) #5536315 reads
length(unique(foram_ord$otu)) #2420 otus
foram_fam <-  foram_df[foram_df$X.Identity.percentage > 96,]
sum(foram_fam$reads) #3525149 reads
length(unique(foram_fam$otu)) #876 otus
foram_sp <-  foram_df[foram_df$X.Identity.percentage > 99.4,]
sum(foram_sp$reads) # 1358462reads
length(unique(foram_sp$otu)) #50 otus
length(unique(foram_sp$Morphospecies)) #27 morphospecies

#check ASV and reads for LBF and SBF
foram_sp %>%
  filter(Large_benthic_foraminifera == "LBF") -> sp_LBF
length(unique(sp_LBF$otu))  #27 otus
sum(sp_LBF$reads) #747873 reads

foram_sp %>%
  filter(Large_benthic_foraminifera == "SBF") -> sp_SBF
length(unique(sp_SBF$otu))  #23 otus
sum(sp_SBF$reads) #610589 reads

###------------------Plot reads per species----------------------------------

#filter species according to LBF and SBF
foram_sp %>%
  group_by(Morphospecies, Large_benthic_foraminifera) %>%
  summarise (nr_reads = sum(reads), .groups = "drop")%>%
  arrange(desc(nr_reads))-> sp_abundance

sp_abundance %>%
  filter(Large_benthic_foraminifera == "LBF") %>%
  pull(Large_benthic_foraminifera) %>%
  unique()-> foram_sp1_LBF
LBF_color <- rep("#481567FF", length(foram_sp1_LBF))

sp_abundance %>%
  filter(Large_benthic_foraminifera == "SBF") %>%
  pull(Large_benthic_foraminifera) %>%
  unique()-> foram_sp1_SBF
SBF_color <- rep("#A3A3A3", length(foram_sp1_SBF))

nested_palette <- c(setNames(SBF_color, foram_sp1_SBF),
                    setNames(LBF_color, foram_sp1_LBF))

#make barplot
ggplot(sp_abundance, aes(x = nr_reads, y = reorder(Morphospecies, nr_reads), fill= Large_benthic_foraminifera )) +
  geom_bar(stat = "identity") +
  geom_text(aes(label = nr_reads), nudge_x = 20000, size = 5) +
  scale_fill_manual(name = "LBF or SBF",
                    values = nested_palette) +
  theme_bw() +
  theme(text = element_text(family = "Arial", face = "plain"),
        axis.text.y =  element_text(size= 16),
        axis.text.x = element_text(angle = 45, hjust = 1, size= 16),
        legend.title = element_text(size = 18, hjust = 0.5),
        legend.text = element_text(size=18),   
        axis.title = element_text(size = 18, face = "bold")) +
  labs(x = "Total number of reads", y = "Species") -> species_level

#ggsave("graphs/species_abundance.svg",        
       plot = species_level,                                              
       device = "svg",                    
       width = 18,  
       height = 8, 
       dpi = 300)
```
**Community composition**


The community composition of foraminifera species is plotted according to each sampling location. The sampling locations are ordered according to DIN:DIP ratio. Different colours are assigned to the families, and the different tones represent the different species within that family. All large benthic foraminifera are giving colours, while the small benthic foraminifera are given greytones. The species are The script to make the composition plot is adapted from https://github.com/EBGirard/COImetabarcoding. 
```{r}
setwd("C:/Users/Venus/OneDrive/Documents/Github/RajaAmpat_forams")

library(colorspace)


foram_sp %>%
  group_by(seqid) %>%
  mutate(tot_abund_seqid = sum(reads)) %>%
  group_by(seqid, Morphospecies) %>%
  mutate(abund_sp = sum(reads)) %>%
  mutate(rel_abund = abund_sp/tot_abund_seqid)->foram_sp1

foram_sp1 <- foram_sp1 %>%
  separate(seqid, 
           into = c("identifier","extraction plate","NBCLAB ID","Year","SED","Repl","Plate","Label","Pos"), 
           sep = "_", remove = FALSE) %>%
  unite("Replicate", Year, SED, Repl, sep = "_", remove = TRUE)

foram_sp1 %>%
  mutate(Location = case_when(
    Location == "Mioskin" ~ "Mioskon",
    Location == "MEHR" ~ "MERHR",
    Location == "EpiCity" ~ "Epaulette City",
    Location == "KaligN" ~ "KaligNorth",
    TRUE ~ Location)) ->foram_sp1

#re-arrange locations according to DIN:DIP ratio
foram_sp1$Location <- factor(foram_sp1$Location, levels = c("Batbitieru","StaffHQ", "MERHR", "LumbaLumba", "Yellit", "South kri",
                                                            "Sabenibnu", "Yenbuba", "Epaulette City", "Villa-MER", "Warakaket", "MER-Bay",
                                                            "Mioskon", "New Reef", "KaligNorth", "KaligRHM", "Sorido Bay",
                                                            "GracesWall","Arborek",  
                                                  "MIS27",  "MIS23", "MIS26","MIS06"))
# assign different colours to families and species
foram_sp1 %>%
  filter(Large_benthic_foraminifera == "LBF") %>%
  filter(Family == "Calcarinidae") %>%
  pull(Morphospecies) %>%
  unique()-> foram_cal_LBF
LBF_cal <- lighten("#0D0087", seq(0, 0.6, length.out = 4))

foram_sp1 %>%
  filter(Large_benthic_foraminifera == "LBF") %>%
  filter(Family == "Nummulitidae") %>%
  pull(Morphospecies) %>%
  unique()-> foram_num_LBF
LBF_num <- lighten("darkmagenta", seq(0, 0.6, length.out = 3))


foram_sp1 %>%
  filter(Large_benthic_foraminifera == "LBF") %>%
  filter(Family == "Peneroplidae") %>%
  pull(Morphospecies) %>%
  unique()-> foram_pen_LBF
LBF_pen <- lighten("#E16462", seq(0, 0.6, length.out = 2))

foram_sp1 %>%
  filter(Large_benthic_foraminifera == "LBF") %>%
  filter(Family == "Amphisteginidae") %>%
  pull(Morphospecies) %>%
  unique()-> foram_amph_LBF
LBF_amph <- darken("#7AD151", seq(0, 0.6, length.out = 4))

foram_sp1 %>%
  filter(Large_benthic_foraminifera == "LBF") %>%
  filter(Family == "Soritidae") %>%
  pull(Morphospecies) %>%
  unique()-> foram_sor_LBF
LBF_sor <- lighten("#FCA636", seq(0, 0.4, length.out = 2))

foram_sp1 %>%
  filter(Large_benthic_foraminifera == "LBF") %>%
  filter(Family == "Alveolinidae") %>%
  pull(Morphospecies) %>%
  unique()-> foram_alv_LBF
LBF_alv <- "#F0F921"

foram_sp1 %>%
  filter(Large_benthic_foraminifera == "SBF") %>%
  pull(Morphospecies) %>%
  unique()-> foram_sp1_SBF
SBF_color <- hcl.colors(11, palette = "LightGrays")

nested_palette <- c(setNames(LBF_alv, foram_alv_LBF),
                    setNames(LBF_amph, foram_amph_LBF),
                    setNames(LBF_cal, foram_cal_LBF),
                    setNames(LBF_num, foram_num_LBF),
                    setNames(LBF_pen, foram_pen_LBF),
                    setNames(LBF_sor, foram_sor_LBF),
                    setNames(SBF_color, foram_sp1_SBF))

foram_sp1 <- foram_sp1 %>%
  mutate(Morphospecies = factor(Morphospecies, levels = names(nested_palette)))


#plot graph
foram_sp1 %>%
  ggplot(aes(fill = factor(Morphospecies), y = rel_abund, x = Replicate)) + 
  geom_bar(position = "fill", stat = "identity") +
  facet_wrap(~ Location, scales = "free_x", ncol = 8)+
  scale_fill_manual(
    name = "Species",
    values = nested_palette,
    guide = guide_legend(title.position = "top")) +
  theme_bw() +
  theme(
    panel.background = element_rect(fill = "white"),
    strip.background = element_blank(),
    text = element_text(family = "Arial"), #segoe ui
    strip.text = element_text(face = "bold", size = 16),
    axis.text.x = element_text(angle = 90, vjust = 0.5, size = 18),
    axis.title.x = element_blank(),
    axis.title.y = element_text(face = "bold", vjust = 0.5, size= 18),
    legend.key.size = unit(0.5, "lines"),
    legend.text = element_text(size = 16),
    legend.title = element_text(size = 18, hjust = 0.5),
    legend.position = "bottom",
    legend.direction = "horizontal"
  ) +
  ylab("Relative abundance of foraminifera species (> 99.4 % ID)") -> com_composition

#ggsave("Graphs/comp_comp.png", plot = com_composition, width = 24, height = 18, dpi = 600)

#ggsave("Graphs/com_comp.svg",        
       plot = com_composition,                                              
       device = "svg",                    
       width = 26, 
       height = 18, 
       dpi = 300)
```
 **Shannon diversity** 
 
 
Now, the Shannon diversity is calculated for each sampling location at ASV-level. The mean Shannon diversity of each replicate measurement is calculated and plotted in a boxplot. Equal variances are calculated with Levene's test, but the variances were not equal, even after log transformation. Therefore, significance is tested with the Kruskal-Wallis test.

```{r}
###------------------------Shannon diversity-------------------------------------
setwd("C:/Users/Venus/OneDrive/Documents/Github/RajaAmpat_forams")

#calculate shannon index per 
foram_df %>%
  group_by(Location, seqid,otu) %>%
  summarise(reads = sum(reads), .groups = "drop") %>% #sum reads per otu per sample
  group_by(Location, seqid) %>%
  mutate(total_abund = sum(reads)) %>%
  mutate(rel_abund = reads/total_abund) %>%
  summarise(shannon = -sum(rel_abund * log(rel_abund)), .groups = "drop") -> foram_shannon

foram_shannon %>%
  group_by(Location) %>%
  summarise(mean_shannon = mean(shannon)) -> foram_shannon_m

foram_shannon$Location <- factor(foram_shannon$Location, levels = c("Batbitieru","StaffHQ", "MEHR", "LumbaLumba", "Yellit", "South kri",
                                                                    "Sabenibnu", "Yenbuba", "EpiCity", "Villa-MER", "Warakaket", "MER-Bay",
                                                                    "Mioskin", "New Reef", "KaligN", "KaligRHM", "Sorido Bay",
                                                                    "GracesWall","Arborek",  
                                                                    "MIS27",  "MIS23", "MIS26","MIS06"))
#plot boxplot
foram_shannon%>%
  ggplot(aes(x= Location, y= shannon)) +
  geom_boxplot() +
  geom_point(size= 1.8, shape = 21, fill = "black")+
  theme_bw() + 
  theme(
    text = element_text(family = "Arial", face = "plain"), 
    axis.text.x = element_text(angle = 90,vjust=0.5, size= 14),
    axis.title.x = element_text(size=16, face = "bold"),
    axis.title.y = element_text(size=16, face = "bold"),
    legend.title  = element_blank(),
    legend.position = "bottom",
    legend.direction = "horizontal") +
  xlab("Location") +
  ylab("Shannon Index") -> shannon_box

#ggsave("Graphs/shannon.svg",        
       plot = shannon_box,                                              
       device = "svg",                    
       width = 10,         
       height = 6, 
       dpi = 300)

#test for signficant difference in shannon diversity
leveneTest(shannon ~ as.factor(Location), data = foram_shannon, center = mean) #variances not equal
foram_shannon$shannon_log <- log(foram_shannon$shannon + 1) #variances still not equal after transformation
kruskal.test(shannon ~ as.factor(Location), data = foram_shannon) #kruskal wallis, p = 0.008497
dunnTest(shannon ~ as.factor(Location), data = foram_shannon, method = "bh")
```
**ASV richness**

Next, ASV richnesss is calculated for each location. It follows a similar code as for the Shannon diversity. 
```{r}
#----------------------ASV richness per sampling location------------------------------------------------
setwd("C:/Users/Venus/OneDrive/Documents/Github/RajaAmpat_forams")

#calculate richness
foram_df %>%
  group_by(seqid, Location)%>%
  summarise(nr_reads = sum(reads),
            nr_otu = n(), .groups = "drop") -> foram_otu

foram_otu %>%
  group_by(Location) %>%
  summarise(richness = mean(nr_otu)) -> foram_otu_m



foram_otu$Location <- factor(foram_otu$Location, levels = c("Batbitieru","StaffHQ", "MEHR", "LumbaLumba", "Yellit", "South kri",
                                                            "Sabenibnu", "Yenbuba", "EpiCity", "Villa-MER", "Warakaket", "MER-Bay",
                                                            "Mioskin", "New Reef", "KaligN", "KaligRHM", "Sorido Bay",
                                                            "GracesWall","Arborek",  
                                                            "MIS27",  "MIS23", "MIS26","MIS06"))

#boxplot
foram_otu%>%
  ggplot(aes(x= Location, y= nr_otu)) +
  geom_boxplot() +
  geom_point(size= 1.8, shape = 21, fill = "black")+
  theme_bw() +
  theme(
    text = element_text(family = "Arial", face = "plain"), 
    axis.text.x = element_text(angle = 90,vjust=0.5, size= 14),
    axis.title.x = element_text(size =16, face = "bold"),
    axis.title.y = element_text(size=16, face = "bold"),
    legend.title  = element_blank(),
    legend.position = "bottom",
    legend.direction = "horizontal") +
  xlab("Location") +
  ylab("Number of ASVs") -> richness_box

#ggsave("Graphs/richness_box.svg",        
       plot = richness_box,                                              
       device = "svg",                    
       width = 10,               
       height = 6, 
       dpi = 300)

#test for signficant difference in richness diversity
leveneTest(nr_otu ~ as.factor(Location), data = foram_otu, center = mean) #variances not equal
foram_otu$nr_otu_log <- log(foram_otu$nr_otu + 1) #variances still not equal after transformation
kruskal.test(nr_otu ~ as.factor(Location), data = foram_otu) #kruskal wallis, p = 0.00005
dunnTest(nr_otu ~ as.factor(Location), data = foram_otu, method = "bh")
```
