---
title: "dbRDA"
author: Venus van Hoof
output: html_notebook
---

This script contains the code to make a dbrda. 

**rda reef**

A rda is made for the reef and the marine lake locations each. ASV abundance data is Hellinger transformed and a rda model is made, as it has a better fit to the data compared to a cca. With forward selection (using the ordi2step function), the best model is selected. This model removes temperature as a predictor, leaving DIN and chlorophyll-a. Continuing with these environmental variables, significance for the mode, its axes and predictors is tested with an anova. Afterwards, data is extracted for the position of the commmunities in the dbRDA plot. Also the positions of significant identified species are extracted. The data is visualized in a dbrda plot.
```{r}
setwd("C:/Users/Venus/OneDrive/Documents/Github/RajaAmpat_forams")

library(tidyverse)
library(vegan)
library(ggh4x)
library(extrafont)
library(ggrepel)

envir <- read.csv("Rdata/clean_envir_datanew.csv", row.names = 1)


###-----------------db RDA for ASVs--------------------------------------------------------

df <- read.csv("Rdata/working_dataset.csv")

foram_df <- subset(df, df$Phylum == "Foraminifera")

envir %>%
  rename(chlor_a = mean_chlora) %>%
  rename(temperature = mean_temp) %>%
  mutate(Location = case_when(
    Location == "MERVilla" ~ "Villa-MER",
    Location == "MERBay" ~"MER-Bay",
    Location == "SoridoBay" ~ "Sorido Bay",
    Location == "NewReef" ~ "New Reef",
    Location == "MiosKon" ~ "Mioskin",
    Location == "KaligNorth" ~ "KaligN",
    Location == "SouthKri" ~ "South kri",
    Location == "WarakaketN" ~ "Warakaket",
    Location == "MERHR" ~ "MEHR",
    Location == "MERStaffHQ" ~"StaffHQ",
    TRUE ~ Location)) -> envir1

#make an ASV table for the reef locations
foram_df %>%
  filter(!Location %in% c("MIS23", "MIS26", "MIS27", "MIS06", "Batbitieru", "Villa-MER")) %>%
  dplyr::select(otu, seqid,reads) %>%
  pivot_wider(names_from = otu, values_from = reads, values_fill = list(reads = 0)) %>%
  as.data.frame()-> foram_df1

rownames(foram_df1) <- foram_df1$seqid
foram_df1 <- foram_df1[, -1]
foram_df1 <- as.matrix(foram_df1)

#hellinger tranformation
foram_hell <- decostand(foram_df1, method = "hellinger")

#add metadata to ASV table
foram_df %>%
  dplyr::select(seqid, Location, Habitat, Benthic.group) %>%
  unique()-> foram_metadat

foram_hell %>%
  as.data.frame() %>%
  rownames_to_column(var = "seqid")%>%
  left_join(foram_metadat, by = "seqid") %>%
  left_join(envir1 %>%
              dplyr::select(-Habitat), by = "Location") -> foram_hell_md

#calculate rda model
foram.rda <- dbrda(foram_hell ~ DIN + temperature + chlor_a, data=foram_hell_md, distance = "bray")
RsquareAdj(foram.rda) #R squared:  0.02399365

#calculate cca model
foram.cca<- cca(foram_hell ~ DIN + temperature + chlor_a, data=foram_hell_md, distance = "bray")
RsquareAdj(foram.cca) # R aquared: 0.009528649 => rda is better fit

#continue with rda model
summary(foram.rda)
#first axis explains 0.4782 ~ 47.82%
#second axis  explains 0.2996 ~29.96%
anova.cca(foram.rda, permutations = 999)
#p = 0.001
anova.cca(foram.rda, by = "axis")
#dbRDA1    1   0.7908 2.3717  0.004 **
#dbRDA2    1   0.4864 1.4586  0.102   
#dbRDA3    1   0.3461 1.0380  0.364  
anova.cca(foram.rda, by = "terms")
#DIN          1   0.5404 1.6206  0.012 * 
#temperature  1   0.3955 1.1861  0.160   
#chlor_a      1   0.6874 2.0617  0.002 **

set.seed(2000)
#forward model selection
model_sel <- ordiR2step(rda(foram_hell ~ 1, data = foram_hell_md[,2627:ncol(foram_hell_md)]), 
                        scope = formula(foram.rda),
                        direction = "forward", Pin = 0.05, R2scope = TRUE) 
model_sel$anova
# DIN           0.0075023  1 -15.037 1.5745  0.022 *
# chlor_a       0.0119291  1 -14.414 1.3360  0.040 *

#----------------continue with only significant variables (DIN & chlor-a)------------------------
foram.rda.r <- dbrda(foram_hell ~ DIN + chlor_a, data=foram_hell_md, distance = "bray")
RsquareAdj(foram.rda.r) #R squared: 0.01910011

#continue with rda model
summary(foram.rda.r)
#first axis  explains 0.6524 ~ 65%
#second axis  explains 0.3476 ~35%
set.seed(2000)
anova.cca(foram.rda.r, permutations = 999) 
#p = 0.002
set.seed(2000)
anova.cca(foram.rda.r, by = "axis")
#dbRDA1    1   0.7608 2.2702  0.002 **
#dbRDA2    1   0.4054 1.2097  0.146 
set.seed(2000)
anova.cca(foram.rda.r, by = "terms")
#DIN       1   0.5404 1.6125  0.015 **
#chlor_a   1   0.6258 1.8674  0.003 ** 

varpart(foram_hell, ~DIN, ~chlor_a, data = foram_hell_md)
#[a+c] = X1            1   0.02056       0.00750     TRUE
#[b+c] = X2            1   0.01905       0.00597     TRUE
#[a+b+c] = X1+X2       2   0.03793       0.01193     TRUE

#extract RDA site scores as a data frame
scores(foram.rda.r, display = "sites") %>% 
  as.data.frame() -> rda.sites.r
rda.sites.r$seqid <- rownames(rda.sites.r)

#merge site scores with metadata
rda.sites.r %>%
  left_join(foram_metadat, by = "seqid") -> rda.sites.r.md

#calculate weighted RDA scores
total_abund <- colSums(foram_hell)
asv_table_nonzero <- foram_hell[,total_abund > 0 ] #includes all ASVs with abundance >0
total_abund_nonzero <- total_abund[total_abund > 0] #only includes total abundance >0
asv_table_nonzero <- asv_table_nonzero[rda.sites.r$seqid,] #keep only rows whose names match rda$seqid

weighted_RDA <- as.matrix(t(asv_table_nonzero)) %*% as.matrix(rda.sites.r[, c("dbRDA1","dbRDA2")])
weighted_RDA1 <- weighted_RDA[, "dbRDA1"] / total_abund_nonzero #normalize
weighted_RDA2 <- weighted_RDA[, "dbRDA2"] / total_abund_nonzero

asv.scores.r <- data.frame(
  RDA1 = weighted_RDA1,
  RDA2 = weighted_RDA2,
  Abundance = total_abund_nonzero,
  ASV = colnames(asv_table_nonzero)
)

#Filter only complete cases for envfit
envir.var.r <- foram_hell_md[,c(1,2627:ncol(foram_hell_md))]
envir.var.r %>% dplyr::select(-temperature) -> envir.var.r
rda.fit.samples.r <- rda.sites.r %>% filter(!is.na(dbRDA1) & !is.na(dbRDA2)) # all samples are maintained

#calculate significant environmental vectors in ordination space
fit.r <- envfit(rda.fit.samples.r[, c("dbRDA1", "dbRDA2")], envir.var.r, permutations = 9999, na.rm=TRUE)
sig.vars.r <- as.data.frame(fit.r$vectors$arrows * sqrt(fit.r$vectors$r))
sig.vars.r$pval <- fit.r$vectors$pvals
sig.vars.r <- sig.vars.r[sig.vars.r$pval <= 0.05, ] #significant vectors


# Save all envfit vectors
envfit_results <- as.data.frame(fit.r$vectors$arrows)
envfit_results$r <- fit.r$vectors$r
envfit_results$r2 <- fit.r$vectors$r^2
envfit_results$pval <- fit.r$vectors$pvals
write.csv(envfit_results, "Rdata/envfit_results_reef_asv.csv", row.names = TRUE)

# Fit asv data to the dbRDA ordination
species.fit.r <- envfit(foram.rda.r, foram_hell, permutations = 999)

sig.sp.r <- as.data.frame(species.fit.r$vectors$arrows * sqrt(species.fit.r$vectors$r))
sig.sp.r$pval <- species.fit.r$vectors$pvals
sig.sp.r <- sig.sp.r[sig.sp.r$pval <= 0.001, ]
row.names(sig.sp.r) -> sig.sp.r$otu
write.csv(sig.sp.r, "Rdata/sig_sp_rda_reef.csv", row.names = TRUE)

#sig.sp.r <-read.csv("data analysis/data/sig_sp_results_reef.csv")
# extract family names
foram_sp <-  foram_df[foram_df$X.Identity.percentage > 99.4,]

foram_sp %>%
  filter(!Location %in% c("MIS23", "MIS26", "MIS27", "MIS06", "Batbitieru", "Villa-MER")) %>%
  dplyr::select(Morphospecies, otu,seqid) %>%
  unique()-> sp_dat_r

# keep  signficant species 
foram_df1 %>%
  as.data.frame() %>%
  rownames_to_column("seqid") %>%
  pivot_longer(-seqid, names_to = "otu", values_to = "nr_reads")-> foram_abund_r

sp_dat_r%>% left_join(foram_abund_r, by = c("otu", "seqid")) -> combdat_r

sig.sp.r %>% left_join(combdat_r, by = "otu") %>%
  drop_na()-> sig.species.r

sig.species.r %>%group_by(Morphospecies) %>%
  summarise(dbRDA1 = weighted.mean(dbRDA1, nr_reads),
            dbRDA2 = weighted.mean(dbRDA2, nr_reads)) -> sig.species

arrow_multiplier <- 3
# Get constrained eigenvalues 
eig.constr.r <- eigenvals(foram.rda.r, model = "constrained")

# Compute axis labels with percentages
x_lab <- paste0("dbRDA1 (", round(eig.constr.r[1] / sum(eig.constr.r) * 100, 1), "%)")
y_lab <- paste0("dbRDA2 (", round(eig.constr.r[2] / sum(eig.constr.r) * 100, 1), "%)")


##ggplot rda per benthic group
ggplot() +
  # geom_point(data = asv.scores.r, aes(x = RDA1, y = RDA2, size = Abundance), color = "grey80", alpha = 0.6) +
  guides(size = "none")+
  geom_point(data = rda.sites.r.md, aes(x = dbRDA1, y = dbRDA2,color = Benthic.group),
             size = 2, alpha = 0.9) +
  scale_color_manual(values = c("hard coral dominated" = "#4B0055", "macro-algae" = "#008298",
                                "macro-algae & hard coral" = "#274983", "rubble - soft coral - encrusting coral" = "#7ED357",
                                "soft coral dominated" = "#FDE333"))+
  stat_ellipse(data = rda.sites.r.md, aes(x = dbRDA1, y =dbRDA2, color = Benthic.group),
               linetype = "dashed", level = 0.95) +
  geom_segment(data = as.data.frame(sig.vars.r),
               aes(x = 0, y = 0, xend = dbRDA1 * arrow_multiplier, yend = dbRDA2 * arrow_multiplier),
               arrow = arrow(length = unit(0.3, "cm")), color = "black") +
  geom_text(data = as.data.frame(sig.vars.r),
            aes(x = dbRDA1 * arrow_multiplier * 1.1, y = dbRDA2 * arrow_multiplier * 1.1,
                label = rownames(sig.vars.r)),
            color = "black", size = 5) +
  geom_text(data = sig.species,
            aes(x = dbRDA1 * arrow_multiplier * 1.1, y = dbRDA2 * arrow_multiplier * 1.1,
                label = Morphospecies),
            color = "darkblue", size = 5) +
  labs(x = x_lab, y = y_lab) +
  ggtitle("Reef") +
  labs(color = "Benthic group dominance") +
  theme_bw() +
  theme(
    legend.key.size = unit(0.5, "lines"),
    legend.text = element_text(size = 14),
    legend.title = element_text(size = 14, hjust = 0.5, face = "bold"),
    legend.position = "bottom",
    legend.direction = "horizontal",
    legend.box = "vertical",
    text = element_text(family = "Arial", face= "plain"),
    axis.text.x = element_text(hjust = 0.5, size = 14),
    axis.text.y = element_text(size = 14),
    axis.title.x = element_text(size = 16, face = "bold"),
    axis.title.y = element_text(size = 16, face = "bold"),
    plot.title = element_text(size = 16, face = "bold"),
    panel.border = element_rect(colour = "black", fill = NA),
    panel.grid = element_blank()) +
  guides(color = guide_legend(title.position = "top")) -> rda_reef_asv

ggsave("Graphs/rda_reefasv.svg",        
       plot = rda_reef_asv,                                              
       device = "svg",                    
       width = 12,                                                                                     
       height = 6, 
       dpi = 300)
```

**rda marine lake**

The same code as above is applied to the marine lake dataset.

```{r}
setwd("C:/Users/Venus/OneDrive/Documents/Github/RajaAmpat_forams")

###-----------------db RDA for marine lake ASVs--------------------------------------------------------

#make an ASV table for the marine lake locations
foram_df %>%
  filter(Location %in% c("MIS23", "MIS26", "MIS27", "MIS06")) %>%
  dplyr::select(otu, seqid,reads) %>%
  pivot_wider(names_from = otu, values_from = reads, values_fill = list(reads = 0)) %>%
  as.data.frame()-> foram_ml

rownames(foram_ml) <- foram_ml$seqid
foram_ml <- foram_ml[, -1]
foram_ml <- as.matrix(foram_ml)

#hellinger tranformation
foram_hellml <- decostand(foram_ml, method = "hellinger")

#add metadata to ASV table
foram_df %>%
  dplyr::select(seqid, Location, Habitat, Benthic.group) %>%
  filter(Location %in% c("MIS23", "MIS26", "MIS27", "MIS06")) %>%
  unique()-> foram_metadat_ml

foram_hellml %>%
  as.data.frame() %>%
  rownames_to_column(var = "seqid")%>%
  left_join(foram_metadat_ml, by = "seqid") %>%
  left_join(envir1 %>%
              dplyr::select(-Habitat), by = "Location") -> foram_hellml_md

#calculate rda model
foram.rda.ml <- dbrda(foram_hellml ~ DIN + temperature + chlor_a, data=foram_hellml_md, distance = "bray")
RsquareAdj(foram.rda.ml) #R squared: 0.3763533

#calculate cca model
foram.cca.ml<- cca(foram_hellml ~ DIN + temperature + chlor_a, data=foram_hellml_md, distance = "bray")
RsquareAdj(foram.cca.ml) # R squared: 0.1190348 => rda is better fit

#continue with rda model
summary(foram.rda.ml)
#first axis  explains 0.4939 ~ 49%
#second axis  explains 0.3247 ~32%

anova.cca(foram.rda.ml, permutations = 999) #p = 0.001
anova.cca(foram.rda.ml, by = "axis")
#dbRDA1    1  0.88229 6.8462  0.001 ***
#dbRDA2    1  0.58012 4.5015  0.001 ***
#dbRDA3    1  0.32410 2.5148  0.001 *** 
anova.cca(foram.rda.ml, by = "terms")
#DIN          1  0.60600 4.7023  0.001 ***
#temperature  1  0.34971 2.7136  0.005 ** 
#chlor_a      1  0.83079 6.4466  0.001 ***

varpart(foram_hellml, ~DIN, ~chlor_a, data = foram_hellml_md)
#[a+c] = X1            1   0.15043       0.10046     TRUE
#[b+c] = X2            1   0.15194       0.10205     TRUE
#[a+b+c] = X1+X2       2   0.28333       0.19374     TRUE

set.seed(2000)
model_selml <- ordiR2step(rda(foram_hellml ~ 1, data = foram_hellml_md[,771:ncol(foram_hellml_md)]), 
                        scope = formula(foram.rda.ml),
                        direction = "forward", R2scope = TRUE) 

model_selml$anova #chlor-a and DIN are significant

#-------------------continue only with only significant variables---------------
#calculate rda model
foram.rda.ml <- dbrda(foram_hellml ~ DIN + chlor_a, data=foram_hellml_md, distance = "bray")
RsquareAdj(foram.rda.ml) #R squared: 0.2012836

summary(foram.rda.ml)
#first axis  explains 0.5619 ~ 56%
#second axis  explains 0.4381 ~44%

set.seed(2000)
anova.cca(foram.rda.ml, permutations = 999) #p = 0.001
set.seed(2000)
anova.cca(foram.rda.ml, by = "axis")
#dbRDA1    1  0.60623 3.6730  0.004 **
#dbRDA2    1  0.47257 2.8632  0.004 **
anova.cca(foram.rda.ml, by = "terms")
#DIN       1  0.60600 3.6716  0.001 ***
#chlor_a   1  0.47279 2.8645  0.001 ***

#extract RDA site scores as a data frame
scores(foram.rda.ml, display = "sites") %>% 
  as.data.frame() -> rda.sites.ml
rda.sites.ml$seqid <- rownames(rda.sites.ml)

#merge site scores with metadata
rda.sites.ml %>%
  left_join(foram_metadat_ml, by = "seqid") -> rda.sites.ml.md

#calculate weighted RDA scores
total_abund.ml <- colSums(foram_hellml)
asv_table_nonzero.ml <- foram_hellml[,total_abund.ml > 0 ] #includes all ASVs with abundance >0
total_abund_nonzero.ml <- total_abund.ml[total_abund.ml > 0] #only includes total abundance >0
asv_table_nonzero.ml <- asv_table_nonzero.ml[rda.sites.ml$seqid,] #keep only rows whose names match rda$seqid

weighted_RDA.ml <- as.matrix(t(asv_table_nonzero.ml)) %*% as.matrix(rda.sites.ml[, c("dbRDA1","dbRDA2")])
weighted_RDA1.ml <- weighted_RDA.ml[, "dbRDA1"] / total_abund_nonzero.ml #normalize
weighted_RDA2.ml <- weighted_RDA.ml[, "dbRDA2"] / total_abund_nonzero.ml

asv.scores.ml <- data.frame(
  RDA1 = weighted_RDA1.ml,
  RDA2 = weighted_RDA2.ml,
  Abundance = total_abund_nonzero.ml,
  ASV = colnames(asv_table_nonzero.ml)
)

#Filter only complete cases for envfit
envir.var.ml <- foram_hellml_md[,c(1,771:772)]
rda.fit.samples.ml <- rda.sites.ml %>% filter(!is.na(dbRDA1) & !is.na(dbRDA2)) # all samples are maintained


#calculate significant environmental vectors in ordination space
fit.ml <- envfit(rda.fit.samples.ml[, c("dbRDA1", "dbRDA2")], envir.var.ml, permutations = 9999, na.rm=TRUE)
sig.vars.ml <- as.data.frame(fit.ml$vectors$arrows * sqrt(fit.ml$vectors$r))
sig.vars.ml$pval <- fit.ml$vectors$pvals 
sig.vars.ml <- sig.vars.ml[sig.vars.ml$pval <= 0.05, ] #significant vectors


# Save all envfit vectors
envfit_results.ml <- as.data.frame(fit.ml$vectors$arrows)
envfit_results.ml$r <- fit.ml$vectors$r
envfit_results.ml$r2 <- fit.ml$vectors$r^2
envfit_results.ml$pval <- fit.ml$vectors$pvals

# Fit asv data to the dbRDA ordination
species.fit.ml <- envfit(foram.rda.ml, foram_hellml, permutations = 999)

sig.sp.ml <- as.data.frame(species.fit.ml$vectors$arrows * sqrt(species.fit.ml$vectors$r))
sig.sp.ml$pval <- species.fit.ml$vectors$pvals
sig.sp.ml <- sig.sp.ml[sig.sp.ml$pval <= 0.001, ]
row.names(sig.sp.ml) -> sig.sp.ml$otu
write.csv(sig.sp.ml, "Rdata/sig_sp_rda_ml.csv", row.names = TRUE)

foram_sp <-  foram_df[foram_df$X.Identity.percentage > 99.4,]
foram_sp %>%
  filter(Location %in% c("MIS23", "MIS26", "MIS27", "MIS06")) %>%
  dplyr::select(Morphospecies, otu,seqid) %>%
  unique()-> sp_dat_ml

# keep  signficant families 
foram_ml %>%
  as.data.frame() %>%
  rownames_to_column("seqid") %>%
  pivot_longer(-seqid, names_to = "otu", values_to = "nr_reads")-> foram_abund_ml

sp_dat_ml%>% left_join(foram_abund_ml, by = c("otu", "seqid")) -> combdat_ml

sig.sp.ml %>% left_join(combdat_ml, by = "otu") %>%
  drop_na()-> sig.species.ml

sig.species.ml %>% group_by(Morphospecies) %>%
  summarise(dbRDA1 = weighted.mean(dbRDA1, nr_reads),
            dbRDA2 = weighted.mean(dbRDA2, nr_reads)) -> sig.spml

arrow_multiplier <- 1
# Get constrained eigenvalues 
eig.constr.ml <- eigenvals(foram.rda.ml, model = "constrained")

# Compute axis labels with percentages
x_lab <- paste0("dbRDA1 (", round(eig.constr.ml[1] / sum(eig.constr.ml) * 100, 1), "%)")
y_lab <- paste0("dbRDA2 (", round(eig.constr.ml[2] / sum(eig.constr.ml) * 100, 1), "%)")


##ggplot rda per benthic group
ggplot() +
  #geom_point(data = asv.scores.ml, aes(x = RDA1, y = RDA2, size = Abundance), color = "grey80", alpha = 0.6) +
  guides(size = "none")+
  geom_point(data = rda.sites.ml.md, aes(x = dbRDA1, y = dbRDA2,color = Benthic.group),
             size = 2, alpha = 0.9) +
  scale_color_manual(values = c("macro-algae & hard coral" = "#274983",
                                "CCA & Sponge" = "#00B28A", "rubble - soft coral - encrusting coral" = "#7ED357")) +
  stat_ellipse(data = rda.sites.ml.md, aes(x = dbRDA1, y =dbRDA2, color = Benthic.group),
               linetype = "dashed", level = 0.95) +
  geom_segment(data = as.data.frame(sig.vars.ml),
               aes(x = 0, y = 0, xend = dbRDA1 * arrow_multiplier, yend = dbRDA2 * arrow_multiplier),
               arrow = arrow(length = unit(0.3, "cm")), color = "black") +
  geom_text(data = as.data.frame(sig.vars.ml),
            aes(x = dbRDA1 * arrow_multiplier * 1.1, y = dbRDA2 * arrow_multiplier * 1.1,
                label = rownames(sig.vars.ml)),
            color = "black", size = 5) +
  geom_text(data = sig.spml,
            aes(x = dbRDA1 * arrow_multiplier * 1.1, y = dbRDA2 * arrow_multiplier * 1.1,
                label = Morphospecies),
            color = "darkblue", size = 5) +
  labs(x = x_lab, y = y_lab) +
  ggtitle("Marine Lake") +
  labs(color = "Benthic group dominance") +
  theme_bw() +
  theme(
    legend.key.size = unit(0.5, "lines"),
    legend.text = element_text(size = 14),
    legend.title = element_text(size = 14, hjust = 0.5, face = "bold"),
    legend.position = "bottom",
    legend.direction = "horizontal",
    legend.box = "vertical",
    text = element_text(family = "Arial", face= "plain"),
    axis.text.x = element_text(hjust = 0.5, size = 14),
    axis.text.y = element_text(size = 14),
    axis.title.x = element_text(size = 16, face = "bold"),
    axis.title.y = element_text(size = 16, face = "bold"),
    plot.title = element_text(size = 16, face = "bold"),
    panel.border = element_rect(colour = "black", fill = NA),
    panel.grid = element_blank()) +
  guides(color = guide_legend(title.position = "top")) -> rda_ml_asv

ggsave("Graphs/rda_mlasv.svg",        
       plot = rda_ml_asv,                                              
       device = "svg",                    
       width = 12,                                                                                     
       height = 6, 
       dpi = 300)
```