---
title: "R Notebook"
author: "Venus van Hoof"
output: html_notebook
---

This Rscript makes a barplot to compare the foraminifera listed in the COI reference database to the foraminifera that were found in literature and in the sediment subsamples. 

The first plot is created to compare the entire foraminifera dataset. The barplot shows the number of orders, families, genera and species that were identified; the overlap with the reference database is shown by slanted lines. For this, two subsets of the data are made: one containing all the identfied species for each source (COI reference database, reviewed studies, or foram subset), while the second subset only contains the number of foraminifera that are shared with the reference database. These subsets are plotter on top each other in ggplot. 
The plot is saved in a SVG file, so it can be accessed for editing later on. 

```{r}
setwd("C:/Users/Venus/OneDrive/Documents/Github/RajaAmpat_forams")

library(tidyverse)
library(readxl)
library(extrafont)
library(colorBlindness)
library(ggpattern)

dat <- read_excel("Rdata/comparison reference database - literature - samples.xlsx", sheet = "Foram_overview")
safe_colorblind_palette <- c("#481567FF", "#238A8DFF","#73D055FF",  "#FDE725FF")


###----------------------- Barplot of foraminifera comparison (literature, database, subsamples)----------------------

#data wrangling
dat %>%
  dplyr::select(source, species, genus, family, order) %>%
  pivot_longer(cols = species:order,
               names_to = "taxonomic_level",
               values_to = "count") %>%
  filter(!is.na(count)) -> dat_long

dat_long %>%
  filter(source%in% c("overlap", "overlap_lit_db", "overlap_sed_db")) -> dat_overlap

# make a separate dataset for the overlap with the COI reference database
dat_overlap %>%
  rename(overlap = source) %>%
  mutate(source = case_when(
    overlap == "overlap" ~ "COI_reference_database",
    overlap  ==  "overlap_lit_db" ~ "literature",
    overlap ==  "overlap_sed_db" ~ "sediment_samples")) %>%
  mutate(taxonomic_level = factor(taxonomic_level, levels = c("order", "family", "genus", "species")))-> dat_overlap

# make a separate dataset for all foraminifera counts
dat_long %>%
  filter(!source %in% c("overlap", "overlap_lit_db", "overlap_sed_db")) %>%
  mutate(taxonomic_level = factor(taxonomic_level, levels = c("order", "family", "genus", "species"))) -> dat_long

#make barplot
ggplot() +
  geom_col(data = dat_long,
           aes(x = taxonomic_level, y = count, fill = taxonomic_level),
           color = "black") + 
  geom_col_pattern(data = dat_overlap, 
                   aes(x = taxonomic_level, y = count, pattern_fill = taxonomic_level),
                   pattern = "stripe",
                   pattern_density= 0.4,
                   pattern_spacing = 0.04,
                   pattern_angle = 30,
                   fill = "white", 
                   pattern_key_scale_factor =0.5, colour = "black") +
  scale_pattern_fill_manual(
    name= "Overlap with reference database",
    values = safe_colorblind_palette) +
  geom_text(data = dat_long,
            aes(x = taxonomic_level, y = count, label =count), #displays label above the bar
            vjust = -0.5, size = 3, inherit.aes = FALSE) + #move label slightly above the bar
  facet_wrap(~source, labeller = as_labeller(c( #multiple plots per plotgroups
    COI_reference_database = "Foraminifera COI \n reference database",
    literature = "Foraminifera literature",
    sediment_samples = "Foraminifera \n sediment subsamples"))) +
  scale_fill_manual(
    name   = "Taxonomic level",
    values = safe_colorblind_palette) +
  labs(x = "Taxonomic Level", y = "Count") +
  ggtitle("All foraminifera") +
  theme_bw() +
  theme(
    text = element_text(family = "Arial", face = "bold", color = "black"),
    plot.title = element_text(hjust=0.5,size =14, face = "bold" ),
    axis.title = element_text(size = 12, face = "bold"),
    axis.text.x = element_text(angle = 45, hjust = 1, size= 10, family = "Arial", face = "plain"),
    strip.text = element_text(size = 10, face = "plain"),
    legend.position = "right") +
  guides(fill = guide_legend(order = 1),
         alpha = guide_legend(order = 2)) +
  coord_cartesian(clip = "off") -> p1 

#ggsave("Graphs/foram_comparison.svg",        
       plot = p1,                                              
       device = "svg",                    
       width = 8,                                                                         
       height = 6, 
       dpi = 300)

```

Next, the same plot is created, but this time only for the large benthic foraminifera.The code remains the same. 

```{r}
LBF <- read_excel("Rdata/comparison reference database - literature - samples.xlsx", sheet = "LBF_overview")

#data wrangling
LBF %>%
  dplyr::select(source, species, genus, family, order) %>%
  pivot_longer(cols = species:order,
               names_to = "taxonomic_level",
               values_to = "count") %>%
  filter(!is.na(count)) -> LBF_long

LBF_long %>%
  filter(source%in% c("overlap", "overlap_lit_db", "overlap_sed_db")) -> LBF_overlap

# make a separate dataset for the overlap with the COI reference database
LBF_overlap %>%
  rename(overlap = source) %>%
  mutate(source = case_when(
    overlap == "overlap" ~ "COI_reference_database",
    overlap  ==  "overlap_lit_db" ~ "literature",
    overlap ==  "overlap_sed_db" ~ "sediment_samples")) %>%
  mutate(taxonomic_level = factor(taxonomic_level, levels = c("order", "family", "genus", "species")))-> LBF_overlap

# make a separate dataset for all foraminifera counts
LBF_long %>%
  filter(!source %in% c("overlap", "overlap_lit_db", "overlap_sed_db")) %>%
  mutate(taxonomic_level = factor(taxonomic_level, levels = c("order", "family", "genus", "species"))) -> LBF_long

# make barplot
ggplot() +
  geom_col(data = LBF_long,
           aes(x = taxonomic_level, y = count, fill = taxonomic_level),
           color = "black") + 
  geom_col_pattern(data =LBF_overlap, 
                   aes(x = taxonomic_level, y = count, pattern_fill = taxonomic_level),
                   pattern = "stripe",
                   pattern_density= 0.4,
                   pattern_spacing = 0.04,
                   pattern_angle = 30,
                   fill = "white", 
                   pattern_key_scale_factor =0.5, colour = "black") +
  scale_pattern_fill_manual(
    name= "Overlap with reference database",
    values = safe_colorblind_palette) +
  geom_text(data = LBF_long,
            aes(x = taxonomic_level, y = count, label =count), #displays label above the bar
            vjust = -0.5, size = 3, inherit.aes = FALSE) + #move label slightly above the bar
  facet_wrap(~source, labeller = as_labeller(c( #multiple lots per plotgroups, labeller renames each facet title
    COI_reference_database = "Foraminifera COI \n reference database",
    literature = "Foraminifera literature",
    sediment_samples = "Foraminifera \n sediment subsamples"
  ))) +
  scale_fill_manual(
    name   = "Taxonomic level",
    values = safe_colorblind_palette
  ) +
  labs(x = "Taxonomic Level", y = "Count") +
  ggtitle("Large benthic foraminifera") +
  theme_bw() +
  theme(
    text = element_text(family = "Arial", face = "bold", color = "black"),
    plot.title = element_text(hjust=0.5,size =14, face = "bold" ),
    axis.title = element_text(size = 12, face = "bold"),
    axis.text.x = element_text(angle = 45, hjust = 1, size= 10, family = "Arial", face = "plain"),
    strip.text = element_text(size = 10, face = "plain"),
    legend.position = "right"
  ) +
  guides(fill = guide_legend(order = 1),
         alpha = guide_legend(order = 2)) +
  coord_cartesian(clip = "off") -> p2 

ggsave("Graphs/LBF_comparison.svg",        
       plot = p2,                                              
       device = "svg",                    
       width = 8,                                                                                  
       height = 6, 
       dpi = 300)


```
