---
title: "NMDS"
author: Venus van Hoof
output: html_notebook
---

**NMDS**

The code executes a nmds and nmds plot using the vegan package. The ASV abundance matrix is hellinger transformed and distances are calculated based on Bray-curtis distance. Data from all sampling locations are included. Different colours are given from sampling locations from reefs and marine lakes. 

```{r}
setwd("C:/Users/Venus/OneDrive/Documents/Github/RajaAmpat_forams")

library(tidyverse)
library(vegan)
library(extrafont)
library(fitdistrplus)
library(ggh4x)
library(car)
library(FSA)
library(pairwiseAdonis)

df <- read.csv("Rdata/working_dataset.csv")
foram_df <- subset(df, df$Phylum == "Foraminifera")
foram_sp <-  foram_df[foram_df$X.Identity.percentage > 99.4,]

###----------------------------NMDS plot for marine lakes and reefs (ASV)------------------------------------
#make ASV table matrix
foram_df %>%
  dplyr::select(otu, seqid,reads) %>%
  pivot_wider(names_from = otu, values_from = reads, values_fill = list(reads = 0)) %>%
  as.data.frame()-> foram_df1
dim(foram_df1) #106 samples and 2981 ASVs

rownames(foram_df1) <- foram_df1$seqid
foram_df1 <- foram_df1[, -1]
foram_df1 <- as.matrix(foram_df1)

#hellinger transformation
foram_hell <- decostand(foram_df1, method = "hellinger") 

#calculate nmds
set.seed(1)
nmds <- metaMDS(foram_hell, distance= "bray", trymax=100, maxit = 1000, k=2) #stress = 0.192083
stressplot(nmds)

#add metadata 
foram_df %>%
  dplyr::select(seqid, Location, Habitat, Benthic.group) %>%
  unique()-> foram_metadat

#plot nmds
scores(nmds, display = "sites") %>%
  as_tibble(rownames = "seqid") %>%
  inner_join(foram_metadat, by = "seqid") %>%
  ggplot(aes(x = NMDS1, y = NMDS2, color = Habitat)) +
  geom_point(size= 3) + 
  scale_color_manual(values = c("Reef" = "#39568CFF", "Marine Lake" = "#55C667FF")) +
  theme_bw() + 
  theme(text = element_text(family = "Arial", face = "plain"),
        legend.text = element_text(size = 14),
        legend.title = element_text(size = 14),
        axis.title = element_text(size = 16, face = "bold", hjust = 0.5)) -> nmds_all

ggsave("Graphs/nmds_all.svg",        
       plot = nmds_all,                                              
       device = "svg",                    
       width = 12,                       
       height = 6, 
       dpi = 300)

adonis2(foram_hell~Habitat,data=foram_metadat,method="bray")
```

Because the reefs are signficantly different from the marine lakes, different nmds plots are made for the reefs and marine lakes.

```{r}
setwd("C:/Users/Venus/OneDrive/Documents/Github/RajaAmpat_forams")


###----------------------------NMDS plot reefs (ASV)--------------------------------------

#make ASV table matrix
foram_df %>%
  filter(!Location %in% c("MIS23", "MIS26", "MIS27", "MIS06")) %>%
  dplyr::select(otu, seqid,reads) %>%
  pivot_wider(names_from = otu, values_from = reads, values_fill = list(reads = 0)) %>%
  as.data.frame()-> foram_reef
dim(foram_reef) #87 samples and 2741 ASVs

rownames(foram_reef) <- foram_reef$seqid
foram_reef <- foram_reef[, -1]
foram_reef <- as.matrix(foram_reef)

#hellinger transformation
foram_hellr <- decostand(foram_reef, method = "hellinger") 

#calculate nmds
set.seed(1)
nmds <- metaMDS(foram_hellr, distance= "bray", trymax=100, maxit = 1000, k=2) #stress = 0.1895536
stressplot(nmds)

#add metadata
foram_df %>%
  filter(!Location %in% c("MIS23", "MIS26", "MIS27", "MIS06")) %>%
  dplyr::select(seqid, Location, Habitat, Benthic.group) %>%
  unique()-> foram_metadatr

#calculate significant species
r.species <- envfit(nmds, foram_hellr)
r.species.plot <- as.data.frame(scores(nmds)$species)
r.species.plot$otu <- rownames(r.species.plot)
r.species.plot <- cbind(r.species.plot, pval=r.species$vectors$pvals)
r.species.sig <- subset(r.species.plot, pval<=0.001)
write.csv(r.species.sig, "Rdata/sig_sp_nmds_reef.csv", row.names = TRUE)

# extract species names
foram_sp <-  foram_df[foram_df$X.Identity.percentage > 99.4,]
foram_sp %>%
  dplyr::select(Morphospecies, otu, Large_benthic_foraminifera) %>%
  unique()-> sp_dat

#get asv abundance data
foram_reef %>%
  as.data.frame() %>%
  rownames_to_column("ID") %>%
  pivot_longer(-ID, names_to = "otu", values_to = "nr_reads") -> reef_abund

#join species names to signficant asvs
r.species.sig %>%
  left_join(reef_abund, by = "otu") %>%
  inner_join(sp_dat, by = "otu") -> r.sp.sig

#collapse species to weighted mean
r.sp.sig %>%group_by(Morphospecies) %>%
  summarise(NMDS1 = weighted.mean(NMDS1, nr_reads),
            NMDS2 = weighted.mean(NMDS2, nr_reads)) %>%
  drop_na()-> sig.species

#plot nmds
scores(nmds, display = "sites") %>%
  as_tibble(rownames = "seqid") %>%
  inner_join(foram_metadat, by = "seqid") %>%
  ggplot(aes(x = NMDS1, y = NMDS2, color = Benthic.group)) +
  geom_point(size= 3) + 
  scale_color_manual(values = c("hard coral dominated" = "#4B0055", "macro-algae" = "#008298",
                                "macro-algae & hard coral" = "#274983", "rubble - soft coral - encrusting coral" = "#7ED357",
                                "soft coral dominated" = "#FDE333")) +
  stat_ellipse(aes(x = NMDS1, y = NMDS2, color = Benthic.group),
               linetype = "dashed", level = 0.95)+ 
  ggrepel::geom_text_repel(data=sig.species, inherit.aes = FALSE,
                           aes(NMDS1, NMDS2, label=Morphospecies),  max.overlaps = 50,
                           color = "black", size = 5)+
  ggtitle("Reef") +
  labs(color = "Benthic group dominance") +
  theme_bw() + 
  theme(text = element_text(family = "Arial", face = "plain"),
        plot.title = element_text(size = 18),
        legend.text = element_text(size = 14),
        legend.title = element_text(size = 14),
        axis.title = element_text(size = 16, face = "bold")) -> nmds_reef_asv

ggsave("Graphs/nmds_reefasv.svg",        
       plot = nmds_reef_asv,                                              
       device = "svg",                    
       width = 12,                                        
       height = 6, 
       dpi = 300)

adonis2(foram_hellr ~Benthic.group,data=foram_metadatr,method="bray")
set.seed(123)
results <- pairwise.adonis(vegdist(foram_hellr,"bray"), foram_metadatr[,4], p.adjust.m = "bonferroni")

library(writexl)
write_xlsx(results, "Rdata/pairwise_nmdsreef.xlsx")
```
Now, the nmds for the marine lakes is calculated.

```{r}
setwd("C:/Users/Venus/OneDrive/Documents/Github/RajaAmpat_forams")

###----------------------------NMDS plot marine lakes (ASV)--------------------------------------

#make ASV table matrix
foram_df %>%
  filter(Location %in% c("MIS23", "MIS26", "MIS27", "MIS06")) %>%
  dplyr::select(otu, seqid,reads) %>%
  pivot_wider(names_from = otu, values_from = reads, values_fill = list(reads = 0)) %>%
  as.data.frame()-> foram_ml
dim(foram_ml) #19 samples and 763 ASVs

rownames(foram_ml) <- foram_ml$seqid
foram_ml <- foram_ml[, -1]
foram_ml <- as.matrix(foram_ml)

#hellinger transformation
foram_hellml <- decostand(foram_ml, method = "hellinger") 

#calculate nmds
set.seed(1)
nmdsml <- metaMDS(foram_hellml, distance= "bray", trymax=100, maxit = 1000, k=2) #stress = 0.1155291
stressplot(nmds)

#add metadata
foram_df %>%
  filter(Location %in% c("MIS23", "MIS26", "MIS27", "MIS06")) %>%
  dplyr::select(seqid, Location, Habitat, Benthic.group) %>%
  unique()-> foram_metadatml

#calculate significant species
ml.species <- envfit(nmdsml, foram_hellml)
ml.species.plot <- as.data.frame(scores(nmdsml)$species)
ml.species.plot$otu <- rownames(ml.species.plot)
ml.species.plot <- cbind(ml.species.plot, pval=ml.species$vectors$pvals)
ml.species.sig <- subset(ml.species.plot, pval<=0.001)
write.csv(ml.species.sig, "Rdata/sig_sp_ml_reef.csv", row.names = TRUE)

# extract family names
foram_sp <-  foram_df[foram_df$X.Identity.percentage > 99.4,]
foram_sp %>%
  dplyr::select(Morphospecies, otu, Large_benthic_foraminifera) %>%
  unique()-> sp_dat

#get asv abundance data
foram_ml %>%
  as.data.frame() %>%
  rownames_to_column("ID") %>%
  pivot_longer(-ID, names_to = "otu", values_to = "nr_reads") -> ml_abund

#join family names to signficant asvs
ml.species.sig %>%
  left_join(ml_abund, by = "otu") %>%
  left_join(sp_dat, by = "otu")-> ml.sp.sig

#collapse families to weighted mean
ml.sp.sig %>%group_by(Morphospecies) %>%
  summarise(NMDS1 = weighted.mean(NMDS1, nr_reads),
            NMDS2 = weighted.mean(NMDS2, nr_reads)) %>%
  drop_na()-> ml.sp

#plot nmds
scores(nmdsml, display = "sites") %>%
  as_tibble(rownames = "seqid") %>%
  inner_join(foram_metadat, by = "seqid") %>%
  ggplot(aes(x = NMDS1, y = NMDS2, color = Benthic.group)) +
  geom_point(size= 3) + 
  scale_color_manual(values = c("macro-algae & hard coral" = "#274983",
                                "CCA & Sponge" = "#00B28A", "rubble - soft coral - encrusting coral" = "#7ED357")) +
  stat_ellipse(aes(x = NMDS1, y = NMDS2, color = Benthic.group),
               linetype = "dashed", level = 0.95)+ #95% confidence ellipse
  ggrepel::geom_text_repel(data=ml.sp, inherit.aes = FALSE,
                           aes(NMDS1, NMDS2, label= Morphospecies),  max.overlaps = 50,
                           color = "black", size = 5)+
  ggtitle("Marine Lake") +
  labs(color = "Benthic group dominance") +
  theme_bw() + 
  theme(text = element_text(family = "Arial", face = "plain"),
        plot.title = element_text(size = 18),
        legend.text = element_text(size = 14),
        legend.title = element_text(size = 14),
        axis.title = element_text(size = 16, face = "bold")) -> nmds_ml_asv

ggsave("Graphs/nmds_mlasv.svg",        
       plot = nmds_ml_asv,                                              
       device = "svg",                    
       width = 12,                                                                                       # You can play with the ratio of width and height to get the best outcome for each plot 
       height = 6, 
       dpi = 300)

adonis2(foram_hellml ~Benthic.group,data=foram_metadatml,method="bray")
set.seed(123)
results <- pairwise.adonis(vegdist(foram_hellml,"bray"), foram_metadatml[,4], p.adjust.m = "bonferroni")

write_xlsx(results, "Rdata/pairwise_adonis_resultsml.xlsx")
```
**Shannon diversity**

The Shannon diversity for the different benthic group dominances is calculated. A boxplot for the shannon diversity is made with a seprate graph for the marine lake and reef benthic groups. Signficant differences in benthic group dominances are calculated for each habitat (reef or marine lake), therefore different subsets are made for each habitat type. 

```{r}
setwd("C:/Users/Venus/OneDrive/Documents/Github/RajaAmpat_forams")

#calculate shannon diversity
foram_df %>%
  group_by(Habitat,Benthic.group, seqid,otu) %>%
  summarise(reads = sum(reads), .groups = "drop") %>% #sum reads per otu per sample
  group_by(Habitat,Benthic.group, seqid) %>%
  mutate(total_abund = sum(reads)) %>%
  mutate(rel_abund = reads/total_abund) %>%
  summarise(shannon = -sum(rel_abund * log(rel_abund)), .groups = "drop") %>%
  group_by(Habitat, Benthic.group) %>%
  mutate(mean_shannon = mean(shannon))-> foram_shannon

#create boxplot
foram_shannon%>%
  ggplot(aes(x= Benthic.group, y= shannon)) +
  geom_boxplot() +
  geom_point(size= 1.8, shape = 21, fill = "black")+
  facet_wrap(~Habitat, scales = "free_x") +
  theme_bw() + 
  theme(
    text = element_text(family = "Arial", face = "plain"),
    strip.text = element_text(size = 14),
    axis.text.x = element_text(angle = 90,vjust=0.5, size= 14),
    axis.title.x = element_text(size=16, face = "bold"),
    axis.title.y = element_text(size=16, face = "bold"),
    legend.title  = element_blank(),
    legend.position = "bottom",
    legend.direction = "horizontal") +
  xlab("Benthic group dominance") +
  ylab("Shannon Index") -> benthicg

ggsave("Graphs/shannon_bg.svg",        
       plot = benthicg,                                              
       device = "svg",                    
       width = 14,                                 
       height = 10, 
       dpi = 300)

#test for signficant difference in benthic group dominance per habitat
#calculate for reefs
foram_df %>%
  filter(Habitat %in% "Reef") %>%
  group_by(Benthic.group, seqid,otu) %>%
  summarise(reads = sum(reads), .groups = "drop") %>% #sum reads per otu per sample
  group_by(Benthic.group, seqid) %>%
  mutate(total_abund = sum(reads)) %>%
  mutate(rel_abund = reads/total_abund) %>%
  summarise(shannon = -sum(rel_abund * log(rel_abund)), .groups = "drop") %>%
  group_by( Benthic.group) %>%
  mutate(mean_shannon = mean(shannon))-> foram_shannon_r

leveneTest(shannon ~ as.factor(Benthic.group), data = foram_shannon_r, center = mean) #variances  equal
anova(lm(shannon ~ as.factor(Benthic.group), data = foram_shannon_r)) # p = 0.3

#calculate for marine lake
foram_df %>%
  filter(Habitat %in% "Marine Lake") %>%
  group_by(Benthic.group, seqid,otu) %>%
  summarise(reads = sum(reads), .groups = "drop") %>% #sum reads per otu per sample
  group_by(Benthic.group, seqid) %>%
  mutate(total_abund = sum(reads)) %>%
  mutate(rel_abund = reads/total_abund) %>%
  summarise(shannon = -sum(rel_abund * log(rel_abund)), .groups = "drop") %>%
  group_by( Benthic.group) %>%
  mutate(mean_shannon = mean(shannon))-> foram_shannon_ml

leveneTest(shannon ~ as.factor(Benthic.group), data = foram_shannon_ml, center = mean) #variances not equal
foram_shannon_ml$shannon_log <- log(foram_shannon_ml$shannon + 1) #variances still not equal after transformation
kruskal.test(shannon ~ as.factor(Benthic.group), data = foram_shannon_ml) # p = 0.2265
```

***ASV richness***
Similar to above, the ASV richness is calculated for each benthic group dominance within each habitat type. The results are visualized in a boxplot. 

```{r}
setwd("C:/Users/Venus/OneDrive/Documents/Github/RajaAmpat_forams")
library(agricolae)

###------------------------richness diversity benthic groups-------------------------------

foram_df %>%
  group_by(Habitat, Benthic.group, seqid)%>%
  summarise(nr_otu = n(), .groups = "drop") %>%
  group_by(Habitat, Benthic.group) %>%
  mutate(mean_otu = mean(nr_otu))-> foram_otu

foram_otu%>%
  ggplot(aes(x= Benthic.group, y= nr_otu)) +
  geom_boxplot() +
  geom_point(size= 1.8, shape = 21, fill = "black")+
  facet_wrap(~Habitat, scales = "free_x") +        
  theme_bw() +
  theme(
    text = element_text(family = "Arial", face = "plain"), 
    strip.text = element_text(size = 14),
    axis.text.x = element_text(angle = 90,vjust=0.5, size= 14),
    axis.title.x = element_text(size=16, face = "bold"),
    axis.title.y = element_text(size=16, face = "bold"),
    legend.title  = element_blank(),
    legend.position = "bottom",
    legend.direction = "horizontal") +
  xlab("Benthic group dominances") +
  ylab("Number of ASVs") ->richness_bg

ggsave("Graphs/richness_bg.svg",        
       plot = richness_bg,                                              
       device = "svg",                    
       width = 14,           
       height = 10, 
       dpi = 300)

#test for signficant difference in benthic group dominance per habitat

#calculate for reefs
foram_df %>%
  filter(Habitat %in% "Reef") %>%
  group_by(Benthic.group, seqid)%>%
  summarise(nr_otu = n(), .groups = "drop") -> foram_otu_r

foram_otu_r$Benthic.group <- as.factor(foram_otu_r$Benthic.group)
leveneTest(nr_otu ~ (Benthic.group), data = foram_otu_r, center = mean) #variances  equal
lm.reef <- lm(nr_otu ~ (Benthic.group), data = foram_otu_r)
anova(lm.reef) # p = 0.001736
HSD.test(lm.reef, trt = "Benthic.group", group = FALSE, console = TRUE)

#calculate for marine lakes
foram_df %>%
  filter(Habitat %in% "Marine Lake") %>%
  group_by(Benthic.group, seqid)%>%
  summarise(nr_otu = n(), .groups = "drop") -> foram_otu_ml

foram_otu_ml$Benthic.group <- as.factor(foram_otu_ml$Benthic.group)
leveneTest(nr_otu ~ (Benthic.group), data = foram_otu_ml, center = mean) #variances not  equal
foram_otu_ml$nr_otulog <- log(foram_otu_ml$nr_otu + 1) #variances still not equal after transformation
kruskal.test(nr_otu ~ (Benthic.group), data = foram_otu_ml) # p = 0.2018
```